---
title: "assignment3"
author: "Abdullah Faqih Al Mubarok"
format: html
editor: visual
---

Today's exercises are based on Chapter 3 in Wooldridge (2018). Feel free to post questions either by [mail](mailto:cep@ifro.ku.dk) or the discussion forum [here](https://ucph.padlet.org/carlemilpless/questions-ehvgql4yj517uhb3). I will monitor both during the day and will be available for questions.

Unless stated otherwise, the primary source of data we will be using is from the `wooldridge` package. Remember, you can access them directly in `R` with the following code:

```{r}
#| eval: false
library(wooldridge)
library(tidyverse)
```

> As long as you have installed the `wooldridge` package using the `install.packages()` command.

Furthermore, it may be useful to call forth a description of the data set by running the following:

```{r}
#| eval: false
sleep_data = sleep75
law_salary_data = lawsch85
bwgs_data = bwght
```

## Problem 3.1

The following model is a simplified version of the multiple regression model used by Biddle and Hamermesh (1990) to study the tradeoff between time spent sleeping and working and to look at other factors affecting sleep:

$$ sleep = \beta_0 + \beta_1 totwrk + \beta_2 educ + \beta_3 age + u,$$

where $sleep$ and $totwrk$ (total work) are measured in minutes per week and $age$ and $educ$ are measured in years.

1.  If adults trade off sleep for work, what is the sign of $\beta_1$?

    it will be minus.

2.  What signs do you think $\beta_2$ and $\beta_3$ will have?\
    For the age it will be positive since the older you are the more time of sleep you need. Meanwhile, for the education, it might be minus since the lesser

3.  Using the data in `sleep75`, the estimated equation is \begin{align} \widehat{sleep} &= 3,638.25 - .148 totwrk -11.3 educ + 2.20 age \\
    n &= 706, R^2 = .113
    \end{align} If someone works five more hours per week, by how many minutes is $sleep$ predicted to decrease? Is this a large tradeoff?

4.  Discuss the sign and magnitude of the estimated coefficient on $educ$.

5.  Would you say $totwrk$, $educ$, and $age$ explain much of the variation in sleep? What other factors might affect the time spent sleeping? Are these likely to be correlated with $totwrk$?

## Problem 3.2

The median starting salary for new law school graduates is determined by

```{=tex}
\begin{align}
    \log(salary) = \beta_0 + \beta_1 LSAT + \beta_2 GPA + \beta_3 \log(libvol) + \beta_4 \log(cost) + \beta_5 rank + u,
\end{align}
```
where $LSAT$ is the median $LSAT$ score for the graduating class, $GPA$ is the median college $GPA$ for the class, $libvol$ is the number of volumes in the law school library, $cost$ is the annual cost of attending law school, and $rank$ is a law school ranking (with $rank = 1$ being the best).

1.  Explain why we would expect $\beta_5 < 0$.

2.  What signs do you expect for the other slope parameters? Justify your answers.

3.  Using the data in `lawsch85`, the estimated equation becomes \begin{align}
    \widehat{\log(salary)} &= 8.34 + .0047 LSAT + .248 GPA + .095 \log(libvol) \\
                        &+.038 \log(cost) -.0033 rank \\
                        n&=136, R^2 = .842
    \end{align} What is the predicted *ceteris paribus* difference in salary for schools with a median $GPA$ difference by one point?

4.  Interpret the coefficient on the variable $\log(libvol)$.

5.  Would you say it is better to attend a higher-ranked law school? How much is the difference in ranking of 20 worth in terms of predicted starting salary?

## Supplementary material 3.1: Deriving the OLS estimator

The following videos use a linear algebra version of the ordinary least squares (OLS) estimation procedure. In further econometrics courses, this is a common representation you will encounter. If you remember your matrix multiplication rules, it is a more elegant solution than the $k$ equations with $k$ unknowns presented in section 3.2f of Wooldridge (2018).

## Problem 3.3

For this exercise, you will need to use the data set `bwght`. A problem of interest to health officials (and others) is determining the effects of smoking during pregnancy on infant health. One measure of infant health is birth weight; a birth weight that is too low can put an infant at risk of contracting various illnesses.

Since factors other than cigarette smoking that affect birth weight are likely to be correlated with smoking, we should consider those factors. For example, higher income generally results in access to better prenatal care and better nutrition for the mother. An equation that recognizes this is:

```{=tex}
\begin{align}
bwght = \beta_0 + \beta_1 cigs + \beta_2 faminc + u,
\end{align}
```
where $faminc$ is family income.

1.  What is the most likely sign of $\beta_2$?

2.  Do you think $cigs$ and $faminc$ are likely to be correlated? Explain why the correlation might be positive or negative.

3.  Now, estimate the equation with and without $faminc$, using the data in `bwght` (the data set we also used yesterday). Report the results in equation form, including the sample size and $R^2$. Discuss your results, focusing on whether adding $faminc$ substantially changes the estimated effect of $cigs$ on $bwght$.

## Problem 3.4

Use the data in `hprice1` to estimate the model:

```{=tex}
\begin{align}
price = \beta_0 + \beta_1 sqrft + \beta_2 bdrms + u,
\end{align}
```
where price is the price of the house measured in thousands of dollars.

1.  Estimate the equation and write out the results in equation form.

2.  What is the estimated increase in price for a house with one more bedroom, holding square footage constant?

3.  What is the estimated increase in price for a house with an additional bedroom that is 140 square feet in size? Compare this to your answer in part 2.

4.  What percentage of the variation in price is explained by square footage and number of bedrooms?

5.  The first house in the sample has $sqrft = 2,438$ and $bdrms = 4$. Find the predicted selling price for this house from the OLS regression line.

::: callout-tip
Try taking a look at the `predict()` function.
:::

6.  The actual selling price of the **first** house in the sample was $300,000$ (so $price = 300$). Find the residual for this house. Does it suggest that the buyer underpaid or overpaid for the house?

## Problem 3.5

Use the data set in `wage2` for this problem. As usual, be sure all of the following regressions contain an intercept.

1.  Run a simple regression of $IQ$ on $educ$ to obtain the slope coefficient, say $\widehat{\delta}_1$

2.  Run the simple regression of $\log(wage)$ on $educ$, and obtain the slope coefficient, $\tilde{\beta}_1$.

3.  Run the multiple regression of $\log(wage)$ on $educ$ and $IQ$ to obtain the slope coefficients $\widehat{\beta}_1$ and $\widehat{\beta}_2$, respectively.

4.  Show that $\tilde{\beta}_1 = \widehat{\beta}_1 + \widehat{\beta}_2 \widehat{\delta}_1$. Explain your result.
